Bootstrapping Multi-atlas Hippocampal Segmentation with \mb

### 2013-04-11 paper.Rnw
Step 1 of a significant re-org of the paper structure -- squashing methods

\documentclass{article}
\usepackage[margin=1in]{geometry}
\usepackage{algorithm, color, algpseudocode, amsmath, url, geometry, ctable}
\usepackage[round,authoryear]{natbib}

@@

\newcommand{\mb}{MAGeT-Brain }
\newcommand{\ants}{ANTS }
\newcommand{\animal}{ANIMAL }

@@

In previous work \citep{Chakravarty2012}, we applied \mb to segmentation
of the human striatum, globus pallidus, and thalamus using a single
histologically-derived atlas. The contribution of the present work is to extend
our approach to the human hippocampus and perform a series of experiments to
rigorous validate the method. First, we conduct an extensive cross-validation of
\mb and basic multi-atlas segmentation on a subset of the ADNI1 dataset to
assess the accuracy of \mb under various parameter settings (number of
atlases and templates, registration and fusion methods). With the best
performing parameter configuration discovered above, we estimate \mb
intra-rater reliability by segmenting separately acquired T1 images of the
atlas subjects. For this experiment, we use the Winterburn atlases: digital
hippocampal subfield segmentations of five in-vivo high-resolution (300u
isotropic) T1-weighted MR scans\citep{Winterburn2013}. To validate \mb in a
real world situation, we segment the entire ADNI1 Complete 1Yr dataset and compare
our segmentations to established automated and manual segmentations.
Additionally, to ensure \mb accuracy across disease categories, we also
compare \mb segmentations to manual segmentations of 139 first episode
schizophrenic patients.

@@

\subsection{\mb Algorithm}
In this paper we will be making a distinction between an atlas and a template --
typically these terms are used roughly interchangeably. The term {\it atlas} is
taken to refer to two image volumes: an intensity image ({\it atlas image}) and
a corresponding manual segmentation image ({\it atlas labels}). {\it Template}
refers more generically to any image and corresponding labelling, manual or
computed, when it is {\it used as} as a model in the segmentation of another
image.  The terms {\it atlas library} and {\it template library} mean a set of
such images. Additionally, we will use the terms {\it target} to refer to an
intensity image for which we would like an segmentation.

@@

%%%%%                             Subjects                                 %%%%%

\subsection{Subjects}
Our experiments use three distinct subject datasets.

For a subset of ADNI1 images, labels of the left and right hippocampi are
available (herein refered to as SNT labels).  Semi-automated hippocampal
volumetry was carried out using a commercially available high dimensional brain
mapping tool (Medtronic Surgical Navigation Technologies, Louisville, CO), that
has previously been validated and compared to manual tracing of the hippocampus
\citep{Hsu2002}. Measurement of hippocampal volume is achieved first by placing
manually 22 control points as local landmarks for the hippocampus on the
individual brain MRI data: one landmark at the hippocampal head, one at the
tail, and four per image (i.e., at the superior, inferior, medial and lateral
boundaries) on five equally spaced images perpendicular to the long axis of the
hippocampus. Second, fluid image transformation is used to match the individual
brains to a template brain \citep{Christensen1997}. The pixels corresponding to
the hippocampus are then labeled and counted to obtain volumes. This method of
hippocampal voluming has a documented reliability of an intraclass coefficient
better than .94 \citep{Hsu2002}.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% ADNI1 Complete 1Yr Dataset Demographics Table                                  %
%                                                                             %
<<ADNI1-scr-demographics, echo=F, results="asis",cache=FALSE>>=
# RID < 2000 ensures only ADNI1 patients
yr1 = read.csv("data/ADNI_baseline_volumes/ADNI1_Complete_1Yr_1.5T_11_15_2012.csv")
yr1$VISCODE <- factor(yr1$Visit, levels = c(1,2,3,4), labels=c("bl","m03","m06","m12"))
yr1 = subset(yr1, !grepl("Scaled_2", Description))
yr1 = yr1[,c("RID","VISCODE")]

\subsubsection{Schizophrenia First Episode Patients}
\label{sec:SZFEP-dataset}
All patients were recruited and treated through the Prevention and Early
Intervention Program for Psychoses (PEPP-Montreal), a specialized early
intervention service at the Douglas Mental Health University Institute in
Montreal, Canada. People aged 15–30 years from the local catchment area
suffering from either affective or non-affective psychosis who had not taken
antipsychotic medication for more than one month with an IQ above 70 were
consecutively admitted as either in- or out-patients. Of those treated at PEPP,
only patients aged 18 to 30 years with no previous history of neurological
disease or head trauma causing loss of consciousness were eligible for the
neuroimaging study; only those suffering from schizophrenia spectrum disorders
were considered for this analysis.  For complete program details see Malla et
al. \citep{Malla2003}.

The hippocampus were traced following a validated protocol developed by Dr Jens
Pruessner (Pruessner et al., 2000). A recent update to this protocol by Dr J
Pruessner in 2006 allows to accurately and consistently subdivide the
hippocampus into three different subregions: head, body, and tail.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% First Episode SZ Demographics Table                                         %
%

<<FEP-SZ-demographics, echo=F, results="asis",cache=FALSE>>=
fep_sz_demographics = read.csv("data/SZ_demographics.csv", na.strings=c("","."))
fep_sz_demographics = subset(fep_sz_demographics, Anon <= 81)
tab <- summary(DX ~ Age + Gender + Handedness + Education + SES + FSIQ,
               data=fep_sz_demographics, method = "reverse", test = FALSE)
latex(tab, file="", ctable=FALSE,
      caption="Schizophrenia First Episode Patient Demographics",
      title="tab:SZFEP-Demographics")
@
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%                       Registration Methods                           %%%%%

\subsection{Registration Methods}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% ANIMAL parameters                                                           %
%                                                                             %
\begin{table}[!tbp]
\begin{center}
\caption{\animal  registration parameters}
\begin{tabular}{l | c c c}
\hline 
Parameters        & Stage 1 & Stage 2 & Stage 3  \tabularnewline
\hline 
Model Blur (FWHM) &   8     &    8    &   4      \tabularnewline
Input Blur (FWHM) &   8     &    8    &   4      \tabularnewline
Iterations        &   30    &    30   &   10     \tabularnewline
Step              &   8x8x8 &  4x4x4  &   2x2x2  \tabularnewline
Sub-Lattice       &   6     &    6    &   6      \tabularnewline
Lattice Diameter  &24x24x24 &12x12x12 & 6x6x6    \tabularnewline
\hline
\end{tabular}
\end{center}
\label{tab:ANIMAL-params}
\end{table}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

%%%%%                           Label Fusion                               %%%%%

\subsection{Label Fusion}

Label fusion is a term given to the process of combining the information from
several candidate labellings for an intensity image into a single labelling.  In
this paper we explore three fusion methods.

\subsubsection{Voxel-wise Majority Vote}

Labels are propagated from all template library images to a target.  Each
output voxel is given the most frequent label at that voxel location amongst
all candidate labellings.  Ties are broken arbitrarily.

\subsubsection{Cross-correlation Weighted Majority Vote}

An optimal combination of targets from the template library has previously been
shown to improve segmentation accuracy \citep{Aljabar2009,Collins2010}.  In this
method, each template library image is ranked in similarity to each unlabelled 
image by the normalized cross-correlation (CC) of image intensities after linear
registration, over a region of interest (ROI) generously encompassing the 
hippocampus.  Only the top ranked template library image labels are used in a
voxel-wise majority vote. The ROI is heuristically defined as the extent of all
atlas labels after linear registration to the template, dilated by three voxels
\citep{Chakravarty2012}.  The number of top ranked template library image labels
is a configurable parameter.

@@

\subsubsection{Normalised Mutual Information Weighted Majority Vote}

This method is similar to cross-correlation weighted voting except that image
similarity is calculated by the normalised mutual information score over the
region of interest\citep{Studholme2001}.  The {\tt itk\_similarity} utility from
the EZMinc toolkit\footnote{\url{https://github.com/vfonov/EZminc}} is used to
calculate the normalised mutual information measure between to images.

%%%%%                        Evaluation Measure                           %%%%%

\subsection{Assessing Segmentation Agreement}

@@

%%%%%
%%%%%                            Experiments                               %%%%%
%%%%%
\subsection{Experiments}

The following experiments were performed to assess the performance of \mb with
various parameter settings as well as on diverse datasets.  In each experiment
we contrast the performance of \mb with that standard single- and multi-atlas
segmentations derived from the same atlas library.

%%%%%
%%%%%                 Experiments: ADNI-1 cross-validationn               %%%%%
%%%%%
\subsubsection{ADNI1 cross-validation}
To test the effect of parameter settings on the \mb algorithm, Monte
Carlo Cross-Validation (MCCV) \citep{Shao1993} was performed using a subset of
the ADNI1 dataset for validation.  This form of cross-validation (described in
detail below) allows us to validate \mb with various atlas and template
library sizes, registration and label fusion methods.

{\bf Dataset evaluated.} 69 1.5T images were arbitrarily selected from the {\em
ADNI1:Complete 1Yr 1.5T} standardized dataset. 23 subjects were chosen from each
disease category (cognitively normal (CN), MCI and AD). Demographics for this
subset are shown in Table \ref{tab:ADNI1-xvalidation-demographics}.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% ADNI1 X-Validation Dataset Demographics Table
%
<<ADNI-Xval-demographics, echo=F, results="asis",cache=FALSE>>=
adni_69_RIDs = c(295,413,619,685,729,782,938,954,1018,1155,907,981,222,324,448,
                 546,553,572,602,610,814,1341,675,681,731,1130,41,68,70,101,
                 249,293,316,414,698,1206,1222,1339,751,1030,3,5,10,16,22,53,
                 168,183,241,1205,328,1095,991,213,159,337,343,642,753,1109,
                 1217,29,56,1188,1293,149,382,431,1202)
tab <- summary(DX.bl ~ AGE + PTGENDER + PTEDUCAT + PTETHCAT + CDRSB + ADAS13 +
               MMSE, data = adnimerge, subset = VISCODE == 'bl' & RID %in%
               adni_69_RIDs, method = "reverse", test=FALSE, overall=TRUE)

latex(tab, file="", size="scriptsize",landscape=FALSE, ctable=FALSE,
      caption="ADNI-1 cross-validation subset demographics", 
      title="tab:ADNI1-xvalidation-demographics")
@
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

{\bf Atlas and template libary.}  The atlas and template libraries are selected
from the dataset evaluated.  Atlases consist of images taken from the
dataset, with corresponding manual labels provided by SNT. Template library 
images were selected at random from remaining images not used as atlases. The
atlas library size ranged from 1 to 9 images. Template library size ranged from
1 to 20 images.

{\bf Registration method.} Both the \ants and \animal registration methods were
used.

{\bf Label fusion.} Majority vote, cross-correlation weighted majority vote, and
Normalized Mutual Information weighted majority vote were used. With the
weighted majority vote fusion methods, the number of top labels used in the
fusion was varied from 1 to 20 images.

{\bf Evaluation.}  Monte Carlo Cross-Validation (MCCV), also known as repeated
random sub-sampling cross-validation, consists of repeated rounds of validation 
in which items from the dataset are randomly sampled (without replacement) and 
assigned to a training set or validation set \cite{Shao1993}.

%%%%%
%%%%%               Experiments: ADNI-1 Complete 1Yr Validation               %%%%%
%%%%%
\subsubsection{ADNI-1 Complete 1Yr Validation}
To test the accuracy of \mb on a real-world task we segment the entire
ADNI-1 dataset using an atlas set that is not representative of the target set.

{\bf Dataset evaluated.} All images from the {\em ADNI1:Complete 1Yr 1.5T}
standardized dataset.

{\bf Atlas and template libary.} The Winterburn atlases were used as the atlas
library.  The template library consisted of 21 randomly selected images from the
ADNI1 data dataset (7 healthy, MCI and AD subjects). Clinical and demographic
data for the template library subjects are shown in Table
\ref{tab:ADNI1-scr-tmpllib-demographics}.

{\bf Registration method.} The \ants non-linear registration algorithm was used,
as it performed best in the ADNI1 cross-validation experiment (see Results).

{\bf Label fusion.} Majority vote was used performed equally well in the ADNI1
cross-validation experiment and is the easiest label fusion method to operate.

{\bf Evaluation.}
Since the hippocampal segmentation protocols differ between the ADNI labels and
Winterburn atlases, this poses a problem for direct evaluation between labels
produced by \mb and the ADNI labels in terms of overlap; we would not expect
different segmentation protocols to have a high degree of overlap.  Instead, to
evaluate the performance of \mb we compare the correlation of \mb segmentation
volumes with manual segmentation (SNT) volumes.  Additionally, we correlate the
hippocampal volumes of established automated segmentation methods to \mb
segmentations.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% ADNI1 Complete 1Yr Template Library Demographics Table
%
<<ADNI1-scr-tmpllib-demographics, echo=F, results="asis",cache=FALSE>>=
scr_tmpllib_RIDs = c(0295, 0413, 0619, 0685, 0729, 0782, 0938, 0954, 1018, 1155,
                 0907, 0981, 0448, 0553, 0814, 1130, 0698, 1339, 1095, 0991,
                 0159)
tab <- summary(DX.bl ~ AGE + PTGENDER + PTEDUCAT + PTETHCAT + CDRSB + ADAS13 +
               MMSE, data = adnimerge, subset = VISCODE == 'bl' & RID %in%
               scr_tmpllib_RIDs, method = "reverse", test=FALSE, overall=TRUE)

latex(tab, landscape=FALSE, ctable=FALSE, file="", size="scriptsize",
      caption="ADNI1 Complete 1Yr Template Library demographics", 
      title="tab:ADNI1-scr-tmpllib-demographics")
@
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%
%%%%%          Experiments: SZ First Episode Patient Validation           %%%%%
%%%%%
\subsubsection{SZ First Episode Patient Validation}
The previous experiments explore \mb on an Alzheimer's disease dataset (ADNI).  To validate
that \mb algorithm works effectively with other diseased brain images, 
the performance of \mb is measured on a Schizophrenia dataset.

{\bf Dataset evaluated.} The entire Schizophrenia First Episode Patient (SZFEP) 
dataset (see section \ref{sec:SZFEP-dataset}).

{\bf Atlas and template libary.} The atlas library is composed of the Winterburn 
T1 atlas images, with whole hippocampus segmentations (no subregions). The 
template library is composed of 21 images selected at random from the SZFEP dataset.

{\bf Registration method.} The \ants registration algorithm is used for image
registration.

{\bf Label fusion.} The majority-vote method is used for label fusion.

{\bf Evaluation.} The manual segmentation protocol used to segment the Winterburn
atlases is similar to, but different from, the protocol used to segment the SZFEP
dataset. Therefore, rather than use an overlap metric, \mb hippocampal volumes
are compared to the corresponding manual segmentation volumes.

%%%%%
%%%%%              Experiments: Winterburn Atlases Validation              %%%%%
%%%%%
\subsubsection{Winterburn Atlases Cross-Validation}
In this experiment, the accuracy of the \mb algorithm on hippocampal subregion
segmentation is tested using a leave-one-out cross-validation (LOOCV) design.

>>>>

\documentclass{article}
\usepackage[margin=1in]{geometry}
\usepackage{algorithm, color, algpseudocode, amsmath, url, geometry, ctable}
\usepackage{titlesec}
\usepackage[round,authoryear]{natbib}

@@

% add \subsubsubsection
\setcounter{secnumdepth}{4}  
\titleformat{\paragraph}
{\normalfont\normalsize\bfseries}{\theparagraph}{1em}{}
\titlespacing*{\paragraph}
{0pt}{3.25ex plus 1ex minus .2ex}{1.5ex plus .2ex}
\newcommand{\subsubsubsection}[1]{\paragraph{#1}}

% shortcuts 
\newcommand{\mb}{MAGeT-Brain }
\newcommand{\ants}{ANTS }
\newcommand{\animal}{ANIMAL }

@@

In previous work \citep{Chakravarty2012}, we applied \mb to segmentation
of the human striatum, globus pallidus, and thalamus using a single
histologically-derived atlas. The contribution of the present work is to extend
our approach to the human hippocampus and perform a series of experiments to
rigorous validate the method. First, we conduct an extensive cross-validation of
\mb and basic multi-atlas segmentation on a subset of the ADNI1 dataset to
assess the accuracy of \mb under various parameter settings (number of
atlases and templates, registration and fusion methods). With the best
performing parameter configuration discovered above, we estimate \mb
intra-rater reliability by segmenting separately acquired T1 images of the
atlas subjects. For this experiment, we use the Winterburn atlases: digital
hippocampal subfield segmentations of five {\em in vivo} high-resolution (300$\mu$
isotropic) T1-weighted MR scans\citep{Winterburn2013}. To validate \mb in a
real world situation, we segment the entire ADNI1 Complete 1Yr dataset and compare
our segmentations to established automated and manual segmentations.
Additionally, to ensure \mb accuracy across disease categories, we also
compare \mb segmentations to manual segmentations of 139 first episode
schizophrenic patients.

@@

\subsection{The \mb Algorithm}
In this paper we will be making a distinction between an atlas and a template --
typically these terms are used roughly interchangeably. The term {\it atlas} is
taken to refer to two image volumes: an intensity image ({\it atlas image}) and
a corresponding manual segmentation image ({\it atlas labels}). {\it Template}
refers more generically to any image and corresponding labelling, manual or
computed, when it is {\it used as} as a model in the segmentation of another
image.  The terms {\it atlas library} and {\it template library} mean a set of
such images. Additionally, we will use the terms {\it target} to refer to an
intensity image for which we would like an segmentation.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%                       Registration Methods                          %%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Registration Methods}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% ANIMAL parameters                    
%                                     
\begin{table}[!tbp]
\begin{center}
\caption{\animal  registration parameters}
\begin{tabular}{l | c c c}
\hline 
Parameters        & Stage 1 & Stage 2 & Stage 3  \tabularnewline
\hline 
Model Blur (FWHM) &   8     &    8    &   4      \tabularnewline
Input Blur (FWHM) &   8     &    8    &   4      \tabularnewline
Iterations        &   30    &    30   &   10     \tabularnewline
Step              &   8x8x8 &  4x4x4  &   2x2x2  \tabularnewline
Sub-Lattice       &   6     &    6    &   6      \tabularnewline
Lattice Diameter  &24x24x24 &12x12x12 & 6x6x6    \tabularnewline
\hline
\end{tabular}
\end{center}
\label{tab:ANIMAL-params}
\end{table}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%                           Label Fusion                              %%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Label fusion methods}
Label fusion is a term given to the process of combining the information from
several candidate labellings for an intensity image into a single labelling.  In
this paper we explore three fusion methods.

\begin{description}
  \item[Voxel-wise Majority Vote] \hfill \\
  Labels are propagated from all template library images to a target.  Each
  output voxel is given the most frequent label at that voxel location amongst
  all candidate labellings.  Ties are broken arbitrarily.

  \item[Cross-correlation Weighted Majority Vote] \hfill \\
  An optimal combination of targets from the template library has previously been
  shown to improve segmentation accuracy \citep{Aljabar2009,Collins2010}.  In this
  method, each template library image is ranked in similarity to each unlabelled 
  image by the normalized cross-correlation (CC) of image intensities after linear
  registration, over a region of interest (ROI) generously encompassing the 
  hippocampus.  Only the top ranked template library image labels are used in a
  voxel-wise majority vote. The ROI is heuristically defined as the extent of all
  atlas labels after linear registration to the template, dilated by three voxels
  \citep{Chakravarty2012}.  The number of top ranked template library image labels
  is a configurable parameter.

@@

  \item[Normalised Mutual Information Weighted Majority Vote] \hfill \\
  This method is similar to cross-correlation weighted voting except that image
  similarity is calculated by the normalised mutual information score over the
  region of interest\citep{Studholme2001}.  The {\tt itk\_similarity} utility
  from the EZMinc toolkit\footnote{\url{https://github.com/vfonov/EZminc}} is
  used to calculate the normalised mutual information measure between to images.
\end{description}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%                            Experiments                              %%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Experiments}

The following section describes the experiments we conducted to assess the
segmentation quality of the \mb algorithm. The first two experiments assess the
validity of \mb using a cross-validation design. Experiment 1 investigates the
accuracy of \mb whole hippocampus segmentation over a wide range of parameter
settings. This enables us to choose the parameter settings offering the best
performance for use in subsequent experiments. Experiment 2 tests hippocampal
subfield segmentation quality. The last two experiments assess the validity of
the \mb algorithm when applied to different diseases: Alzheimer's disease
(Experiment 3) and first episode schizophrenia patients (Experiment 4).

%The ADNI experiment includes comparison of volumes with other automated
%segmentation methods. Evaluation is done by correlating volumes b/c overlap
%metrics assume identical segmentation protocol.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%                                                                     %%%%%
%%%%%         Experiment 1: Whole Hippocampus Cross-Validation            %%%%%
%%%%%                                                                     %%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection*{Experiment 1: Whole Hippocampus Cross-Validation}

To test the effect of parameter settings on the \mb algorithm, Monte
Carlo Cross-Validation (MCCV) \citep{Shao1993} was performed using a subset of
the ADNI1 dataset for validation (see below).  This form of cross-validation
allows us to validate \mb with various atlas and template library sizes,
registration and label fusion methods.

Monte Carlo Cross-Validation (MCCV), also known as repeated random sub-sampling
cross-validation, consists of repeated rounds of validation in which items from
the dataset are randomly sampled (without replacement) and assigned to a
training set or validation set \cite{Shao1993}.

\subsubsubsection{Dataset evaluated.}

69 1.5T images were arbitrarily selected from the {\em ADNI1:Complete 1Yr 1.5T}
standardized dataset. 23 subjects were chosen from each disease category
(cognitively normal (CN), MCI and AD). Demographics for this subset are shown in
Table \ref{tab:ADNI1-xvalidation-demographics}.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
% ADNI1 X-Validation Dataset Demographics Table
%
<<ADNI-Xval-demographics, echo=F, results="asis",cache=FALSE>>=
adni_69_RIDs = c(295,413,619,685,729,782,938,954,1018,1155,907,981,222,324,448,
                 546,553,572,602,610,814,1341,675,681,731,1130,41,68,70,101,
                 249,293,316,414,698,1206,1222,1339,751,1030,3,5,10,16,22,53,
                 168,183,241,1205,328,1095,991,213,159,337,343,642,753,1109,
                 1217,29,56,1188,1293,149,382,431,1202)
tab <- summary(DX.bl ~ AGE + PTGENDER + PTEDUCAT + PTETHCAT + CDRSB + ADAS13 +
               MMSE, data = adnimerge, subset = VISCODE == 'bl' & RID %in%
               adni_69_RIDs, method = "reverse", test=FALSE, overall=TRUE)

latex(tab, file="", size="scriptsize",landscape=FALSE, ctable=FALSE,
      caption="ADNI-1 cross-validation subset demographics", 
      title="tab:ADNI1-xvalidation-demographics")
@
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

The atlas and template libraries are selected from the dataset evaluated.
Atlases consist of images taken from the dataset, with corresponding manual
labels provided by SNT. Template library images were selected at random from
remaining images not used as atlases. The atlas library size ranged from 1 to 9
images. Template library size ranged from 1 to 20 images.

Both the \ants and \animal registration methods were used.

Majority vote, cross-correlation weighted majority vote, and
Normalized Mutual Information weighted majority vote were used. With the
weighted majority vote fusion methods, the number of top labels used in the
fusion was varied from 1 to 20 images.

\subsubsubsection{Evaluation method}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%                                                                     %%%%%
%%%%%        Experiment 2: Hippocampal Subfield Cross-Validation          %%%%%
%%%%%                                                                     %%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection{Winterburn Atlases Cross-Validation}
In this experiment, the accuracy of the \mb algorithm on hippocampal subregion
segmentation is tested using a leave-one-out cross-validation (LOOCV) design.
Leave-one-out-cross-validation (LOOCV) is performed separately for each dataset
evaluated (WA-BRAVO, and WA-Resampled) as follows. Each subject in the dataset
is segmented by \mb using an atlas library in which that subject's image is
excluded.

The atlas library is composed of the Winterburn T1 atlas images and hippocampal
subregion segmentations. The template library consists of 15 3T T1 images
(0.9mm-isotropic voxels) of healthy subjects as well as all of the images from
the dataset being evaluated.

- note registration and fusion method were chosen 
\todo{reference, demographics for health control images}.

\subsubsubsection{Dataset evaluated} 
\mb segmentations are evaluated on two datasets. The WA-BRAVO dataset consists
of separately acquired 3T T1 BRAVO images (0.9mm-isotropic voxel) of four of the
five Winterburn atlas subjects.  The WA-Resampled dataset consists of the
Winterburn atlas images and segmentations downsampled to 0.9mm-isotropic voxel
resolution.

\subsubsubsection{Evaluation method}  
The segmentation volumes of each hippocampal subregion are compared to
the expert manual segmentations of the unmodified Winterburn atlases.  As a
point of comparison, we also calculate the subregion volumes from the Winterburn
atlas segmentations after downsampling to 0.9mm-isotropic voxels.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%                                                                     %%%%%
%%%%%         Experiment 3: Application of \mb to the segmentation        %%%%%
%%%%%                       of schizophrenia patients                     %%%%%
%%%%%                                                                     %%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection{SZ First Episode Patient Validation}
To validate that \mb algorithm works effectively with diseased brain images the
performance of \mb is measured on a Schizophrenia dataset.

The atlas library is composed of the Winterburn T1 atlas images, with whole
hippocampus segmentations (no subregions). The template library is composed of
21 images selected at random from the SZFEP dataset.

- note registration and fusion method were chosen 
\subsubsubsection{Dataset Evaluated}

\subsubsubsection{Evaluation method}
The manual segmentation protocol used to segment the Winterburn atlases is
similar to, but different from, the protocol used to segment the SZFEP dataset.
Therefore, rather than use an overlap metric, \mb hippocampal volumes are
compared to the corresponding manual segmentation volumes.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%                                                                     %%%%%
%%%%%         EXPERIMENT 3: Application of \mb to the segmentation        %%%%%
%%%%%                       of ADNI subjects                              %%%%%
%%%%%                                                                     %%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection{Experiment 4: Segmentation of Alzheimer's Disease Patients}

To test the accuracy of \mb on a real-world task we segment the entire
ADNI-1 dataset using an atlas set that is not representative of the target set.

The Winterburn atlases were used as the atlas
library.  The template library consisted of 21 randomly selected images from the
ADNI1 data dataset (7 healthy, MCI and AD subjects).

- note registration and fusion method were chosen

\ref{tab:ADNI1-scr-tmpllib-demographics}.

\subsubsubsection{Dataset evaluated} 
All images from the {\em ADNI1:Complete 1Yr 1.5T}
standardized dataset.

\subsubsubsection{Evaluation method}
Since the hippocampal segmentation protocols differ between the ADNI labels and
Winterburn atlases, this poses a problem for direct evaluation between labels
produced by \mb and the ADNI labels in terms of overlap; we would not expect
different segmentation protocols to have a high degree of overlap.  Instead, to
evaluate the performance of \mb we compare the correlation of \mb segmentation
volumes with manual segmentation (SNT) volumes.  Additionally, we correlate the
hippocampal volumes of established automated segmentation methods to \mb
segmentations.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
% ADNI1 Complete 1Yr Template Library Demographics Table
%
%
% <<ADNI1-scr-tmpllib-demographics, echo=F, results="asis",cache=FALSE>>=
% scr_tmpllib_RIDs = c(0295, 0413, 0619, 0685, 0729, 0782, 0938, 0954, 1018, 1155,
%                  0907, 0981, 0448, 0553, 0814, 1130, 0698, 1339, 1095, 0991,
%                  0159)
% tab <- summary(DX.bl ~ AGE + PTGENDER + PTEDUCAT + PTETHCAT + CDRSB + ADAS13 +
%                MMSE, data = adnimerge, subset = VISCODE == 'bl' & RID %in%
%                scr_tmpllib_RIDs, method = "reverse", test=FALSE, overall=TRUE)
% 
% latex(tab, landscape=FALSE, ctable=FALSE, file="", size="scriptsize",
%       caption="ADNI1 Complete 1Yr Template Library demographics", 
%       title="tab:ADNI1-scr-tmpllib-demographics")
% @
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%                             Subjects                                 %%%%%
\subsection{Subjects}

\subsubsubsection{FEP dataset}
All patients were recruited and treated through the Prevention and Early
Intervention Program for Psychoses (PEPP-Montreal), a specialized early
intervention service at the Douglas Mental Health University Institute in
Montreal, Canada. People aged 15–30 years from the local catchment area
suffering from either affective or non-affective psychosis who had not taken
antipsychotic medication for more than one month with an IQ above 70 were
consecutively admitted as either in- or out-patients. Of those treated at PEPP,
only patients aged 18 to 30 years with no previous history of neurological
disease or head trauma causing loss of consciousness were eligible for the
neuroimaging study; only those suffering from schizophrenia spectrum disorders
were considered for this analysis.  For complete program details see Malla et
al. \citep{Malla2003}.

The hippocampus were traced following a validated protocol developed by Dr Jens
Pruessner \todo{cite: (Pruessner et al., 2000)}. A recent update to this
protocol by Dr J Pruessner in 2006 allows to accurately and consistently
subdivide the hippocampus into three different subregions: head, body, and tail.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
% First Episode SZ Demographics Table
%
<<FEP-SZ-demographics, echo=F, results="asis",cache=FALSE>>=
fep_sz_demographics = read.csv("data/SZ_demographics.csv", na.strings=c("","."))
fep_sz_demographics = subset(fep_sz_demographics, Anon <= 81)
tab <- summary(DX ~ Age + Gender + Handedness + Education + SES + FSIQ,
               data=fep_sz_demographics, method = "reverse", test = FALSE)
latex(tab, file="", ctable=FALSE,
      caption="Schizophrenia First Episode Patient Demographics",
      title="tab:SZFEP-Demographics")
@
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
